Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2020-05-04 paper-bedpca.tex
need to change format now

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix. These steps includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness. Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these. 
In the following, we outline these pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,zou2010quantification,abdellaoui2013population,prive2017efficient}.
Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic associations within these LD regions \cite[]{zou2010quantification}.  
Second, another issue may arise when projecting a new study dataset to the PCA space computed from a reference dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
This same issue also arises when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This potential issue has been demonstrated for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

@@

Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance.
We use these distances for many applications in this paper. 
When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
If $x$ represents PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust}.
We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
Then, we compute $K$ PC scores and loadings (Section \ref{randomSVD}).
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances of these PC loadings (Section \ref{maha}).  
To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window with a radius of 50 variants by default).

@@

To project a target genotype dataset to the PCA space from a reference dataset (e.g.\ the 1000 genomes data), we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the individuals used for computing PCA, we provide function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).

@@

When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}). 
These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure \ref{fig:UKBB-loadings}) and that was not previously reported in \cite{price2008long}.

@@

As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique. It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the most computationally demanding step when computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).

@@

Once PCA is done, one should check the PC scores (scores of each individual for each PCA dimension) and PC loadings (weights for each variant for each PCA dimension). 
We differentiate PCs in 3 broad types: the ones capturing LD structure, the ones capturing population structure, and noise.
We expect population structure to be evenly distributed along the genome so that loadings are normally distributed around 0 (with small effect sizes).
In contrast, long-range LD structure is essentially capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
As in many applications, we believe a compromise between signal and noise should be preferred. Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that do not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).

>>>>

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt,prive2020performing} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix. These steps include removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness. 
Some genetic analyses may also require to restrict to individuals of homogeneous ancestry.
Many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these. 
In the following, we outline these pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,zou2010quantification,abdellaoui2013population,prive2017efficient}.
Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic associations within these LD regions \cite[]{zou2010quantification}.  
Second, another issue may arise when projecting a new study dataset to the PCA space computed from a reference dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
This same issue also arises when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This potential issue has been demonstrated for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

@@

Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance.
We use these distances for many applications in this paper. 
When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
If $x$ represents PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust}.
We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr, and the direct computation of these robust distances in function \texttt{dist\_ogk}.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
Note that this procedure does not require removing any known long-range LD region a priori.
We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
Then, we compute $K$ PC scores and loadings (Section \ref{randomSVD}).
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances of these PC loadings (Section \ref{maha}).  
To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window with a radius of 50 variants by default).

@@

To project a target genotype dataset to the PCA space from a reference dataset (e.g.\ the 1000 genomes data), we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the individuals used for computing PCA, we provide function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA (cf.\ section \ref{proj-related}).

@@

When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, i.e.\ it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}). 
These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure \ref{fig:UKBB-loadings}) and that was not previously reported in \cite{price2008long}.

@@

As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique. It takes 12 minutes more to perform the OADP projection of all 488,371 UKBB individuals onto the PCA space computed using the 1000G data.
Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the most computationally demanding step when computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).

@@

Once PCA is done, one should check the PC scores (scores of each individual for each PCA dimension) and PC loadings (weights for each variant for each PCA dimension). 
We differentiate PCs in 3 broad types: the ones capturing LD structure, the ones capturing population structure, and noise.
We expect population structure to be evenly distributed along the genome so that loadings are normally distributed around 0 (with small effect sizes).
In contrast, long-range LD structure is essentially capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
PCs with no visible population stratification, i.e.\ where all individuals are normally distributed around 0, can be considered as noise.   
As in many applications, we believe a compromise between signal and noise should be preferred. Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that do not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).
